Manuscript

### 2025-03-25 Manuscript_MD_ScoreReLiability.qmd
change bullet points into text

Across twelve exemplary archival datasets, we demonstrated that claims
made in H&S and W&D do not hold in empirical observations. Even though
removing the impact of score reliability on ES heterogeneity by means of
attenuation correction procedures, they do not lead to a reduction in
heterogeneity as claimed in H&S and W&D. Inferences derived from the
equation supplied in H&S (ref), @eq-HS, also do not warrant such a
conclusion. Alternatively, describing standardised effect sizes in terms
of a ratio distribution and approximating its variance, we demonstrated
that differences in score reliability may both inflate or deflate
variance in ES, depending on how the relative heterogeneity in score
variance changes due to the attenuation correction procedure. The metric
$R_2$, compared with the metric $R_1$ or the average score reliability,
informs us in how far score variance heterogeneity decreases due to the
attenuation correction.

These results fit in with work recently published/preprinted by
Olsson-Collentine et al. (2023). In a large simulation scheme, they
demonstrate that differences in score reliability across administrations
typically deflate heterogeneity in uncorrected correlations. Only as the
true heterogeneity in correlations grows larger, score reliability
differences actually inflate heterogeneity as claimed in W&D. While
correlations and standardized effect sizes are not identical, the way
score (un)reliability affects these parameters is highly similar. Both
correlations and ES are deflated by (un)reliability, implying that an
attenuation correction increases mean values of these parameters, while
the impact on standard deviations used for standardisation purposes
needs to be checked via $R_2$.

The analytical arguments presented indicate that an attenuation
correction procedure, even if no differences between score reliability
exist, lead to larger ES heterogeneity. It appears that, as the
meta-analytical mean ES grows larger due to the correction procedure, so
does the ES heterogeneity, if homogeneity is present. If heterogeneity
is present on the other hand and $R_2 < R_1^3$, still a reduction in ES
heterogeneity is not inevitable. It can be argued that, for the
differences in ES to grow smaller, the reduction in relative
heterogeneity in score variance needs to be substantially large, so the
increase in ES heterogeneity due to the larger mean ES can be
“outpaced”. However, we did not observe this pattern in the 12 phenomena
that allowed for such an analysis.

Heterogeneity in ES even in direct replications, where experimental
factors are held as constant as possible, is already substantial
(Renkewitz, Fünderich, & Beinhauer, 2024). However, so far we have
demonstrated that in several scenarios, differences in score reliability
may in fact be masking, or “deflating” ES heterogeneity. This implies
that the already substantial heterogeneity across direct replications
may very well be even larger, if ES were corrected for (un)reliability.
Understanding heterogeneity as an extent to which a theory or phenomenon
is understood, deflated estimates of heterogeneity imply that the theory
or phenomenon is understood even worse than initially assumed.

The empirical arguments presented are based on a rather small number of
non-representative data-sets. Based on the combination of analytical and
empirical arguments, we are convinced that these differences in score
reliability across replications oftentimes mask true heterogeneity. In
realistic meta-analytic studies of conceptual replications, data stems
from a variety of sources. Across different populations, different
research settings or designs are used with varying or adjusted research
instruments. It is unlikely that, across these replications, we can
expect either the measuring quality or the true score variance to be
stable. However, as demonstrated above, differences in score reliability
and true score variance tend to lead to larger corrected ES
heterogeneity, implying that observed ES heterogeneity was reduced by
differences in score reliability. However, whether these results
actually generalize beyond these data-sets remains to be seen.
Unfortunately, the behavioural sciences are in dire need of open-data
that resemble multi-site replications. As the majority of results
discussed over the last years (in e.g. ManyLabs or Registered
Replication Reports) employ single-indicator scales as dependent
measures, score reliability can not be easily estimated in order to
replicate our analyses.

Paragraph on implications -\> data collected using no specific sampling
schemes - little surprise that the radomness induced by sampling
schemes, which manifests itself in true score variance heterogeneity,
largely surpasses the heterogeneity induced by imperfect measurement.

The analytical arguments presented following Equation 17 rested on the
use of the delta-method, as Equation 17 contained parameters of the
score standard deviations’ distributions, while the $R_1$ and $R_2$
metrics were designed to discuss score variances. The delta-method rests
on a number of assumptions: (1) approximations derived from the
delta-method require for the transformed variable to follow a marginal
normal distribution (ref). Since the variable under discussion here is
the standard deviation, which is bounded to be larger than zero and has
a non-normal sampling distribution (ref), this assumption is most likely
violated; (2) the derivative of the transformation function needs to be
available. As the transformation function is essentially just the square
root, this assumption should hold up; (3) as the delta-method
approximates the parameters of the transformed variable using the
first-order Taylor-expansion, this implies that higher-order terms are
required to be negligible. In practice, this implies that the variance
of the untransformed variable needs to be small. However, as this entire
procedure discusses the influence of variance in standard deviations on
ES heterogeneity, introduced by differences in score reliability, this
assumption is probably violated as well. The violation of these
assumptions likely introduce a bias in estimates derived from these
approximations (ref). However, the inequality derived with use of the
delta-method is not used to actually derive any estimates, but to
discuss how an attenuation correction procedure can affect ES
heterogeneity. Therefore, we believe that the violation of these
assumptions is defensible in this case, as the analytical arguments
derived from the inequality should still hold up, nevertheless.

@@

The work done by Hunter and Schmidt (2014) was crucial in informing and
guiding methodology development in the field of meta-analyses.
Similarly, Wiernik and Dahlke (2020) raise a number of important points
that have been neglected in the application of meta-analytic research
over the last decades. We strongly agree with their ideas that
underappreciated (differences in un-)reliability introduce substantial
biases in meta-analytic estimates and tests in Psychology, which need to
be corrected. Future applications of meta-analyses on standardized ES
need to take score reliability into account to arrive at correct
estimates and inferences. However, the meta-analyst should not expect
that correcting ES for score reliability will reduce the extent of
heterogeneity identified.

>>>>

We have examined meta-analytic heterogeneity in standardized mean
difference. Since score reliability is often discussed as a source for
such heterogeneity, we estimated the heterogeneity before and after
correcting for differences in reliability coefficients using an
attenuation correction procedure. We found that a) in the majority of
cases, the change in ES heterogeneity is small to non-existent. This
implies that, across the data-sets we got to examine, differences in
score reliability do not serve as an appropriate explanation for ES
heterogeneity. We also found that b) in the majority of cases, the
direction of change in ES heterogeneity does not align with assumptions
formulated in W&D and H&S. In almost all cases (every single case, if we
restrict ourself to phenomena where we found statistically significant
differences in ES), the attenuation correction led to increased
estimates of ES heterogeneity. This implies that the claims found in W&D
and H&S stem from a general misunderstanding of how score reliability
can affect ESs.

Using analytical arguments, we derived a broad rule, outlining under
which conditions we can expect an attenuation correction to reduce ES
heterogeneity. In general, score reliability affects ESs by inflating
the score variance, as the (square root of the) score variance is used
to standardize the mean difference. The conditions we found depend on
how both the mean value and the heterogeneity of score variance are
reduced by correcting for imperfect score reliability. We introduced two
novel metrics: $R_1$, describing the fraction of mean true score
variance in the mean total score variance, therefore mirroring the
average score reliability coefficient, and $R_2$, describing the
fraction of true score variance heterogeneity in the total score
variance heterogeneity. We find that, as long as $R_1^3 < R_2$, an
attenuation correction will increase ES heterogeneity.

Since $R_1$ is carrying the power of 3, it implies that the reduction in
mean score variance is weighted more heavily than the reduction in score
variance heterogeneity. This is best described through an example: We
know what an acceptable average score reliability looks like in the area
of Psychology. Researchers largely target a score reliability of at
least .7, preferably .8. An average score reliability of .7 implies that
70% of the variance observed in scores can actually be attributed to
true underlying differences. From our derived rule, we know that, as
long as $R_2$ is larger or equal to $.7^3 = .343$, the attenuation
correction procedure can only increase ES heterogeneity. If we want to
explain ES heterogeneity through score reliability, even though only 30%
of the variance observed in score can be attributed to random
measurement error, at least 65.7%, almost two thirds, of the
heterogeneity in score variance would need to be attributable to
differences in the error score variance. In other words, at most 34.3%
of heterogeneity in score variances may be due to actual differences in
how the latent variable is distributed (in terms of spread) across the
administration sites.

This also means that ES heterogeneity is best reduced, if average score
reliability ($R_1$) is substantially large. In such a scenario, the
$R_2$ is more likely to be low enough for the inequality to no longer
hold. The larger the average score reliability, the more likely it is
that the attenuation correction procedure can actually reduce ES
heterogeneity. However, as the correction would only be rather small in
such cases, it remains to be seen whether the reduction in ES
heterogeneity would be sufficient to claim that differences in score
reliability actually posed a relevant moderator in such a case.

In the examples we re-analysed throughout this text, as well as the
majority of research studies done in the behavioural sciences, no proper
sampling techniques are employed. Most samples where these instruments
are applied come to be through convenience sampling, where students
self-select into studies. Without a valid sampling scheme, we
essentially have have no basis to assume anything about the true score
variance. Most importantly, however, we have no reason to assume that
the true score variance could potentially be homogeneous or stable
across administration sites. Instead, it is very likely that the
convenience sampling procedures produce samples that vary extremely
strongly in how they are put together, leading to large degrees of true
score variance heterogeneity. At the same time, score reliability is a
low-hanging fruit for researchers to optimise most researchers are aware
and trained in assessing score reliability.

Taken together, these facets lead to what we observed throughout this
text: Reliability coefficients are in an acceptable range, leading to a
small, but constant reduction in mean score variance if we apply an
attenuation correction. At the same time, there is substantial variation
in true score variance, implying that most of the score variance
heterogeneity can not be attributed to differences in error score
variance. Since only differences in error score variance can be
corrected for, the attenuation correction has a very small effect on
heterogeneity estimates in score variance. Therefore, as indicated by
our rule, an attenuation correction in these setting can rarely deliver
a reduction in ES heterogeneity. This also means that, in such settings,
score reliability does not explain any ES heterogeneity. If anything, it
masks it to a small degree.

If we, in the space of behavioural sciences, keep making use of
convenience samples simply because they are so convenient, we can not
expect anything to change. Most likely, ES heterogeneity will remain a
substantial factor hampering the interpretability of original and
replication studies. We could learn more about standardized effect sizes
and why our phenomena vary across settings, populations or continents by
not following the current status quo of convenience samples. Until this
status quo changes, however, we will keep wasting valuable resources by
attempts of discussing ES heterogeneity or assessing uninformative
differences in score reliability.

These results fit in with work recently published/preprinted by
Olsson-Collentine et al. (2023). In a large simulation scheme, they
demonstrate that differences in score reliability across administrations
typically deflate heterogeneity in uncorrected correlations. Only as the
true heterogeneity in correlations grows larger, score reliability
differences actually inflate heterogeneity as claimed in W&D. While
correlations and standardized effect sizes are not identical, the way
score (un)reliability affects these parameters is highly similar.

Heterogeneity in ES even in direct replications, where experimental
factors are held as constant as possible, is already substantial
(Renkewitz, Fünderich, & Beinhauer, 2024). However, so far we have
demonstrated that in several scenarios, differences in score reliability
did not serve as appropriate moderators or explanations of ES
heterogeneity. Across most phenomena, we failed to reduce ES
heterogeneity. If anything, we even increased ES heterogeneity, albeit
to a very small degree. Generally, concerning the multi-site replication
projects or collaborative research projects, accounting for differences
in score reliability did not help us gain a better understanding of the
phenomena.

The empirical arguments presented are based on a rather small number of
non-representative data-sets. Based on the combination of analytical and
empirical arguments, we are convinced that these differences in score
reliability across replications in the current state of Psychology and
the behavioural sciences, do not serve as appropriate explanations of ES
heterogeneity. However, whether these results actually generalize beyond
these data-sets remains to be seen. Unfortunately, the behavioural
sciences are in dire need of open-data that resemble multi-site
replications. As the majority of results discussed over the last years
(in e.g. ManyLabs or Registered Replication Reports) employ
single-indicator scales as dependent measures, score reliability can not
be easily estimated in order to replicate our analyses.

The analytical arguments presented following Equation 17 rested on the
use of the delta-method, as Equation 17 contained parameters of the
score standard deviations’ distributions, while the $R_1$ and $R_2$
metrics were designed to discuss score variances. The delta-method rests
on a number of assumptions: (1) approximations derived from the
delta-method require for the transformed variable to follow a marginal
normal distribution (ref). Since the variable under discussion here is
the standard deviation, which is bounded to be larger than zero and has
a non-normal sampling distribution (ref), this assumption is most likely
violated; (2) the derivative of the transformation function needs to be
available. As the transformation function is essentially just the square
root, this assumption should hold up; (3) as the delta-method
approximates the parameters of the transformed variable using the
first-order Taylor-expansion, this implies that higher-order terms are
required to be negligible. In practice, this implies that the variance
of the untransformed variable needs to be small. However, as this entire
procedure discusses the influence of variance in standard deviations on
ES heterogeneity, introduced by differences in score reliability, this
assumption is probably violated as well. The violation of these
assumptions likely introduce a bias in estimates derived from these
approximations (ref). However, the inequality derived with use of the
delta-method is not used to actually derive any estimates, but to
discuss how an attenuation correction procedure can affect ES
heterogeneity. Therefore, we believe that the violation of these
assumptions is defensible in this case, as the analytical arguments
derived from the inequality should still hold up, nevertheless. \[add to
appendix?\] Additionally, we used the empirical examples available to
assess in how far estimates derived using the delta-method
approximations diverge from estimates derived without it. The
differences identified are rather miniscule at the third decimal,
reinforcing our belief that these assumption violations are defensible.

@@

The work done by Hunter and Schmidt (2014) was crucial in informing and
guiding methodology development in the field of meta-analyses.
Similarly, Wiernik and Dahlke (2020) raise a number of important points
that have been neglected in the application of meta-analytic research
over the last decades. We generally agree with their ideas that
underappreciated (differences in un-)reliability can introduce
substantial biases in meta-analytic estimates and tests in Psychology.
However, based on our findings, we believe that the issues created by
ignoring random sampling is much more dire and warrants immediate
attention. Only once we can be sure that the variance in true scores or
latent factors is actually meaningful for some population, we can turn
towards imprecise measurements, expecting correction procedures to
explain unwarranted heterogeneity. In the current state of Psychology,
meta-analysts should not expect that correcting ES for score reliability
will actually explain heterogeneity identified.
